The large scale clustering in dispersion measure space

### 2015-05-18 writeup/dm_space_distortions.tex
Polished figures.

\documentclass[twocolumn,prl,nofootinbib]{revtex4-1}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{verbatim}
\usepackage{times}
\usepackage{afterpage}

@@

@@

% This paragraph needs to be reorganized.
However, redshift is not the only measure of radial distance. In particular,
dispersion of electromagnetic signals due to intervening plasma
is readily measured from fast radio bursts (FRB) which are a class
of radio transient first reported in \citet{lorimer}. There have since been 
\ldots such bursts detected \citep{all_bursts} and total rate of bursts
been estimated to be \ldots. The dispersion measure
(DM) is a line-of-site integral of the free electron density and can thus be
used as a distance estimate.
If the progenitors of these bursts are at cosmological distances, with the
majority of the dispersion caused by electrons in the inter-galactic medium,
then the average DM-distance relation can be modelled \citep{someone}.

Neither redshift nor dispersion measure are perfect proxies for radial distance.
Redshifts are systematically biased by the peculiar velocities of the object
relative to the Hubble flow. This leads to additional apparent clustering 
in redshift
space which was first calculated by \citet{kaiser} using linear theory.
Similarly, distance estimates from dispersion measure will be biased by
inhomogeneities in the electron density, again leading to additional apparent
clustering of sources in dispersion measure space.  In this letter we calculate
these DM space distortions and consider the detectability of the signal by
upcoming surveys.

@@

We wish to relate the density of a tracer $f$ measured in DM space to its
density in real space. We follow the derivation in \citet{Kaiser} of the
redshift space distortions. Start by noting that the total number of tracers
in a volume element is the same in both spaces:
\begin{equation}
\label{e:density}
n_{fs}(\vec x_s) \ud^3\vec x_s = n_{f}(\vec x) \ud^3\vec x.
\end{equation}
We split the density into a homogeneous part plus perturbations,
\begin{equation}
    \bar{n}_{fs}(\chi_s)\left[ 1 + \delta_{fs}(\vec x_s)\right] \ud^3\vec x_s
    = \bar{n}_{f}(\chi)\left[ 1 + \delta_f(\vec x)\right] \ud^3\vec x.
\end{equation}
Averaged over the sky, the density should be the same in DM 
space as in real space,
\begin{equation}
\bar{n}_{fs}(\chi) = \bar{n}_f(\chi).
\end{equation}
Therefore,
\begin{align}
\bar{n}_{fs}(\chi_s) 
    &= \bar{n}_{f}(\chi) + (\chi_s - \chi)\frac{\ud \bar{n}_f}{\ud \chi}\\
    &= \bar{n}_{f}(\chi)
       + \frac{\ud \bar{n}_f}{\ud \chi}\int_0^\chi \ud \chi' \delta_e(\hat n \chi')
       \label{e:nfs}
\end{align}

@@

For brevity in the following sections we define the coefficient of the integral
term as
\begin{equation}
    A(\chi) \equiv \frac{1}{\bar{n}_f}\frac{\ud \bar{n}_f}{\ud \chi}
    + \frac{2}{\chi}.
\end{equation}

@@

Equation~\ref{e:Clss} can be simplified substantially by adopting the small
angle and Limber approximations \citep{extened_limber}.  The small angle
approximation eliminates the $k$ integral over spherical Bessel functions, replacing it
with a Fourier transform, and is valid for $\ell \gg 1$. The Limber
approximation assumes that only modes with small radial component of their wave
vector contribute to the radial integrals and is valid if the power spectra
evolve slowly compared to the correlation length (which has already been
assumed). With these approximations we have
\begin{align}
    &C^{ss}_\ell(\chi, \chi') \approx
    \nonumber \\ & \quad
    \frac{1}{\bar\chi^2}
        \int_{-\infty}^\infty\frac{\ud k_\parallel}{(2 \pi)} 
        e^{i k_\parallel (\chi - \chi')}
        P_{[ee + ff - 2ef]}
        (\sqrt{k_\parallel^2 + \nu^2/\bar\chi^2}, \bar\chi)
    \nonumber \\ & \quad +
    A(\chi) A(\chi')
    \int_0^{\chi_{\rm min}}\ud\chi''
    \frac{1}{\chi^{\prime\prime 2}}
    P_{ee}(\nu/\chi'',\chi'')
    \nonumber \\ & \quad +
    \frac{A(\chi_{\max})}{\chi_{\min}^2}
    P_{[ee - ef]}(\nu/\chi_{\min}, \chi_{\min}),
    \label{e:Clss_limber}
\end{align}
where $\nu\equiv\ell + 1/2$, $\chi_{\min} \equiv \min(\chi, \chi')$,
and $\chi_{\max} \equiv \max(\chi,
\chi')$ .  We've found that these approximations are excellent for $\ell
\gtrsim 10$
and use this form for the remainder of the paper.

@@

There are a number of effects that will complicate the measurement of the
above calculated angular power spectrum by a real survey of dispersion 
measures.  The first is
that the model that the free electron density decomposes into a homogeneous
part plus a random field of perturbations is not correct. Firstly, all signals
must pass through the Milky Way's galactic disk, halo and local environment
which will contribute to the dispersion measure. However, this is only a 
single function of $\hat n$  which can presumably be well
measured. In addition, this function could in principal be well characterized
by independent data such as observations of pulsars.  As such, the local
contribution is unlikely to be a limiting obstacle. \red{cite geansler paper?}

More concerning is the dispersion from the burst progenitor's environment
as well as the halo in which it resides. These electrons will presumably be
clustered around the burst progenitor on scales far smaller than DM space
structure maps and thus are not accounted for in the $\delta_e$--$\delta_f$
correlations included in the above formalism.  This will have a mean
contribution, common to all bursts and which may be epoch dependant, which will
modify the mean DM--distance relation. This will need to be calibrated in some
why or else fit with nuisance parameters during parameter estimation.  There
will also be a stochastic piece which varies from burst to burst.  This will
limit the precision at which $\chi_s$ can be measured, thus limiting the
resolution of DM space density maps in the radial direction.

@@

There is an advantage to the lack of kernel in the DM space integral term.
Lensing tomography---the use of sources at multiple redshifts to unmix
spatial scales---is complicated by the kernel. The kernel's shape depends on
the source redshift, reweighting the line of sight and
making tomography inexact.  In contrast the integrand in the integral
term is completely independent of the limit of integration, meaning
contributions from different parts of the line of sight (and thus different
spatial scales) can be separated exactly. Again, a detailed study of DM space
tomography is beyond the scope of this work.  The avoid our results being
sensitive to the hard-to-model small scales, we assume that the contribution to the
integral term from radial distances $\chi'' < 500\,{\rm MPc}/h$ are well
measured by the correlations with these radial slices. We thus neglect the
information from these slices in our sensitivity measurements and ignore
contributions to the integral term from below this distance. The smallest
spatial scale that contributes the our plotted angular power spectra out to
$\ell=1000$ is thus $k=2\,h/{\rm MPc}$.

@@

\begin{figure}
    \includegraphics[scale=0.42]{"figures/n_f"}
    \caption{
        \label{f:n_f}
        \red{Need to fix normalization of top panel.}
    }
\end{figure}

\begin{figure}
    \includegraphics[scale=0.42]{"figures/terms"}
    \caption{
        \label{f:terms}
        Terms in the DM-space cross-correlation angular power spectrum.
        All terms are evaluated at $\bar\chi=1000\,{\rm MPc}/h$ and 
        $\chi - \chi' = 10\,{\rm MPc}/h$, except the lowest most 
        ``local term'' curve which is for $\chi - \chi'=50\,{\rm MPc}/h$. 
        While
        the other two terms are highly insensitive to the separation of
        the radial slices, the local term drops rapidly at high separations.
    }
\end{figure}

@@

Here we consider the sensitivity of a survey with a total of $10^4$ dispersion
measures from FRBs distributed uniformly over half the sky. This is in rough
approximation to what could be achieved by the Canadian Hydrogen Intensity
Mapping Experiment's now funded FRB instrument over a three year survey
\citep{who}.
To give an idea of the overall sensitivity of the survey, we collapse
$C^{ss}_{\ell,ij}$ to a single function of $\ell$ by taking a weighted average
of all pairs of radial bins. We use bin widths of $100\,{\rm MPc}/h$ over the
range $500\,{\rm MPc}/h$ to $3500\,{\rm MPc}/h$.
For weights in the radial bin average, we use the signal to noise-squared
at $\ell=100$, which maximizes the signal to noise at that $\ell$ and
is near optimal at other multipoles.
We exclude the $i = j$ bin pairs
as they would be contaminated by noise bias in a real survey.
This effectively discards the contribution from the local term. The result is shown
in Figure~\ref{f:sensitivity}.

\begin{figure}
    \includegraphics[scale=0.42]{"figures/sensitivity"}
    \caption{
        \label{f:sensitivity}
        Sensitivity of a survey with $10^4$ dispersion measures distributed
        uniformly over half the sky.  Plotted is the cross-correlation
        power spectrum weighted averaged over all pairs of radial bins. Weights
        are chosen to maximize signal to noise at $\ell=100$.
    }
\end{figure}

@@

\bibliography{dm_space}

>>>>

\documentclass[twocolumn,prl,nofootinbib]{revtex4-1}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{verbatim}
\usepackage{times}
\usepackage{afterpage}
\usepackage{epstopdf}

@@

\input{AAS_macros.tex}

@@

% This paragraph needs to be reorganized.
However, redshift is not the only measure of radial distance. In particular,
dispersion of electromagnetic signals due to intervening plasma
is readily measured from fast radio bursts (FRB) which are a class
of radio transient first reported in \citet{2007Sci...318..777L}.
There have since been 
10 such bursts detected \citep{2012MNRAS.425L..71K,2013Sci...341...53T, 2014ApJ...790..101S, 2014ApJ...792...19B, 2015MNRAS.447..246P,2015ApJ...799L...5R} and total rate of bursts
been estimated to be $3.3^{+5}_{-2.5} \times 10^3$
per sky per day \cite{2015arXiv150500834R}. 
The dispersion measure
(DM) is a line-of-site integral of the free electron density and can thus be
used as a distance estimate.
If the progenitors of these bursts are at cosmological distances, with the
majority of the dispersion caused by electrons in the inter-galactic medium,
then the average DM-distance relation can be modelled
\citep{2004MNRAS.348..999I, 2003ApJ...598L..79I}.

Neither redshift nor dispersion measure are perfect proxies for radial distance.
Redshifts are systematically biased by the peculiar velocities of the object
relative to the Hubble flow. This leads to additional apparent clustering 
in redshift
space which was first calculated by \citet{1987MNRAS.227....1K}
using linear theory.
Similarly, distance estimates from dispersion measure will be biased by
inhomogeneities in the electron density \citep{2014ApJ...780L..33M},
again leading to additional apparent
clustering of sources in dispersion measure space.  In this letter we calculate
these DM space distortions and consider the detectability of the signal by
upcoming surveys.

@@

We wish to relate the density of a tracer $f$ measured in DM space to its
density in real space. We follow the derivation in \citet{1987MNRAS.227....1K}
of the
redshift space distortions. Start by noting that the total number of tracers
in a volume element is the same in both spaces:
\begin{equation}
\label{e:density}
n_{fs}(\vec x_s) \ud^3\vec x_s = n_{f}(\vec x) \ud^3\vec x.
\end{equation}
We split the density into a homogeneous part plus perturbations,
\begin{equation}
    \bar{n}_{fs}(\chi_s)\left[ 1 + \delta_{fs}(\vec x_s)\right] \ud^3\vec x_s
    = \bar{n}_{f}(\chi)\left[ 1 + \delta_f(\vec x)\right] \ud^3\vec x.
\end{equation}
Averaged over the sky, the density should be the same in DM 
space as in real space,
\begin{equation}
\bar{n}_{fs}(\chi) = \bar{n}_f(\chi).
\end{equation}
Therefore,
\begin{align}
\bar{n}_{fs}(\chi_s) 
    &= \bar{n}_{f}(\chi) + (\chi_s - \chi)\frac{\ud \bar{n}_f}{\ud \chi}\\
    &= \bar{n}_{f}(\chi)
       + \frac{\ud \bar{n}_f}{\ud \chi}\int_0^\chi \ud \chi' \delta_e(\hat n \chi')
       \label{e:nfs}
\end{align}

@@

For brevity in the following sections we define the coefficient of the integral
term as
\begin{equation}
    \label{e:A}
    A(\chi) \equiv \frac{1}{\bar{n}_f}\frac{\ud \bar{n}_f}{\ud \chi}
    + \frac{2}{\chi}.
\end{equation}

@@

Equation~\ref{e:Clss} can be simplified substantially by adopting the small
angle and Limber approximations \citep{1953ApJ...117..134L,
1992ApJ...388..272K, 1998ApJ...498...26K}.  The small angle
approximation eliminates the $k$ integral over spherical Bessel functions, replacing it
with a Fourier transform, and is valid for $\ell \gg 1$. The Limber
approximation assumes that only modes with small radial component of their wave
vector contribute to the radial integrals and is valid if the power spectra
evolve slowly compared to the correlation length \citep{2008PhRvD..78l3506L}
(which has already been
assumed). With these approximations we have
\begin{align}
    &C^{ss}_\ell(\chi, \chi') \approx
    \nonumber \\ & \quad
    \frac{1}{\bar\chi^2}
        \int_{-\infty}^\infty\frac{\ud k_\parallel}{(2 \pi)} 
        e^{i k_\parallel (\chi - \chi')}
        P_{[ee + ff - 2ef]}
        (\sqrt{k_\parallel^2 + \nu^2/\bar\chi^2}, \bar\chi)
    \nonumber \\ & \quad +
    A(\chi) A(\chi')
    \int_0^{\chi_{\rm min}}\ud\chi''
    \frac{1}{\chi^{\prime\prime 2}}
    P_{ee}(\nu/\chi'',\chi'')
    \nonumber \\ & \quad +
    \frac{A(\chi_{\max})}{\chi_{\min}^2}
    P_{[ee - ef]}(\nu/\chi_{\min}, \chi_{\min}),
    \label{e:Clss_limber}
\end{align}
where $\nu\equiv\ell + 1/2$, $\chi_{\min} \equiv \min(\chi, \chi')$,
and $\chi_{\max} \equiv \max(\chi,
\chi')$ .  We've found that these approximations are excellent for $\ell
\gtrsim 10$
and use this form for the remainder of the paper.

@@

There are a number of effects that will complicate the measurement of the
above calculated angular power spectrum by a real survey of dispersion 
measures.  The first is
that the model that the free electron density decomposes into a homogeneous
part plus a random field of perturbations is not correct. Firstly, all signals
must pass through the Milky Way's galactic disk, halo and local environment
which will contribute to the dispersion measure. However, this is only a 
single function of $\hat n$  which can presumably be well
measured. In addition, this function could in principal be well characterized
by independent data such as observations of pulsars.  As such, it should be
possible to subtract the local
contribution, as was done in
\citet{2014arXiv1412.4829D}, and this is unlikely to be a limiting obstacle.

More concerning is the dispersion from the burst progenitor's environment
as well as the halo in which it resides. These electrons will presumably be
clustered around the burst progenitor on scales far smaller than DM-space
structure maps and thus are not accounted for in the $\delta_e$--$\delta_f$
correlations included in the above formalism.  This will have a mean
contribution, common to all bursts and which may be epoch dependant, which will
modify the mean DM--distance relation. This will need to be calibrated in some
why or else fit with nuisance parameters during parameter estimation.  There
will also be a stochastic piece which varies from burst to burst.  This will
limit the precision at which $\chi_s$ can be measured, thus limiting the
resolution of DM space density maps in the radial direction.

@@

There is an advantage to the lack of kernel in the DM-space integral term.
Lensing tomography---the use of sources at multiple redshifts to unmix
spatial scales---is complicated by the kernel. The kernel's shape depends on
the source redshift, reweighting the line of sight and
making tomography inexact.  In contrast the integrand in the integral
term is completely independent of the limit of integration, meaning
contributions from different parts of the line of sight (and thus different
spatial scales) can be separated exactly. Again, a detailed study of DM space
tomography is beyond the scope of this work.  The avoid our results being
sensitive to the hard-to-model small scales, we assume that the contribution to the
integral term from radial distances $\chi'' < 500\,{\rm MPc}/h$ are well
measured by the correlations with these radial slices. We thus neglect the
information from these slices in our sensitivity measurements and ignore
contributions to the integral term from below this distance. The smallest
spatial scale that contributes the our plotted angular power spectra out to
$\ell=1000$ is thus $k=2\,h/{\rm MPc}$.

@@

\begin{figure}
    \includegraphics[scale=0.42]{figures/n_f}
    \caption{
        \label{f:n_f}
        \emph{Top:} Comoving density of sources for toy model, as given in
        Equation~\ref{e:n_f}. Normalization is
        such that the survey has a total of $10^4$ FRBs.
        \emph{Bottom:} Coefficient $A(\chi)$ (defined in Equation~\ref{e:A})
        in toy model.
    }
\end{figure}

\begin{figure}
    \includegraphics[scale=0.42]{figures/terms}
    \caption{
        \label{f:terms}
        Terms in the DM-space cross-correlation angular power spectrum.
        All terms are evaluated at $\bar\chi=1000\,{\rm MPc}/h$ and 
        $\chi - \chi' = 10\,{\rm MPc}/h$, except the lowest most 
        ``local term'' curve which is for $\chi - \chi'=50\,{\rm MPc}/h$. 
        While
        the other two terms are highly insensitive to the separation of
        the radial slices, the local term drops rapidly at high separations.
    }
\end{figure}

@@

Here we consider the sensitivity of a survey with a total of $10^4$ dispersion
measures from FRBs distributed uniformly over half the sky. This is in rough
approximation to what could be achieved by the Canadian Hydrogen Intensity
Mapping Experiment's now funded FRB instrument over a three year survey.
To give an idea of the overall sensitivity of the survey, we collapse
$C^{ss}_{\ell,ij}$ to a single function of $\ell$ by taking a weighted average
of all pairs of radial bins. We use bin widths of $100\,{\rm MPc}/h$ over the
range $500\,{\rm MPc}/h$ to $3500\,{\rm MPc}/h$.
For weights in the radial bin average, we use the signal over noise-squared
at $\ell=100$, which maximizes the signal to noise at that $\ell$ and
is near optimal at other multipoles.
We exclude the $i = j$ bin pairs
as they would be contaminated by noise bias in a real survey.
This effectively discards the contribution from the local term. The result is shown
in Figure~\ref{f:sensitivity}.

\begin{figure}
    \includegraphics[scale=0.42]{figures/sensitivity}
    \caption{
        \label{f:sensitivity}
        Sensitivity of a survey with $10^4$ dispersion measures distributed
        uniformly over half the sky.  Plotted is the cross-correlation
        power spectrum weighted averaged over all pairs of radial bins. Weights
        are chosen to maximize signal to noise at $\ell=100$.
    }
\end{figure}

@@

\bibliography{refs}
